---
title: Incorporating Probabilistic Sensitivity Analysis  
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    number_sections: TRUE
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{Incorporating Probabilistic Sensitivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Required R-packages 
```{r, message = FALSE, warning=FALSE, results='hide'}
#install.packages("expm")
#install.packages("hesim")
#install.packages("data.table")
library(expm)
library(hesim)
library(data.table)

rm(list=ls())
```

# Model settings
set starting distribution over the 4 health states
```{r, message=FALSE}
initial<-matrix(c(1,0,0,0),nrow=1)
```
set duration of follow-up
```{r, message=FALSE}
follow_up <- 20
```
set discount rates
```{r, message=FALSE}
discount_LY <- 0.00
discount_cost <- 0.06
```
number of PSA simulations
```{r, message=FALSE}
PSA_samples  <- 1000
```

# Model input
## Transition matrix monotherapy
Create PSA samples for transition matrix monotherapy based on Dirichlet distribution (using rdirichlet_mat function from hesim package)
```{r, message=FALSE}
transitions_mono <- matrix(c(1251, 350, 116, 17,
                             0, 731, 512, 15,
                             0, 0, 1312, 437,
                             0, 0, 0, 469),
                             nrow = 4, byrow = TRUE)
p_mono = rdirichlet_mat(n=PSA_samples,transitions_mono)
```
Show first 10 PSA samples
```{r, message=FALSE}
print(p_mono[, , 1:10])
```

## Relative treatment effect combination therapy
Create PSA samples for relative risk according to normal distribution of log-transformed relative risk
```{r, message=FALSE}
RR<-0.509
RR_low<-0.365
RR_high<-0.710

RR_combi <- exp( rnorm(PSA_samples, log(RR), (log(RR_high)-log(RR_low))/3.92 )   )
```
Show first 10 PSA samples
```{r, message=FALSE}
RR_combi[1:10]
```

## Transition matrix combination therapy
Create PSA samples for transition matrix for combination therapy based on samples for monotherapy and RR
```{r, message=FALSE}
p_combi <- array(NA, dim = c(4, 4, PSA_samples))
for (i in 1:PSA_samples){ 
p_combi[,,i] <- matrix(c((1-p_mono[1,2,i]*RR_combi[i]-p_mono[1,3,i]*RR_combi[i]-p_mono[1,4,i]*RR_combi[i]), p_mono[1,2,i]*RR_combi[i], p_mono[1,3,i]*RR_combi[i], p_mono[1,4,i]*RR_combi[i],
                                      0, (1-p_mono[2,3,i]*RR_combi[i]-p_mono[2,4,i]*RR_combi[i]), p_mono[2,3,i]*RR_combi[i], p_mono[2,4,i]*RR_combi[i],
                                      0, 0, (1-p_mono[3,4,i]*RR_combi[i]), p_mono[3,4,i]*RR_combi[i],
                                      0, 0, 0, 1),
                                      ncol = 4, nrow = 4, byrow = TRUE)
}
```
Show first 10 PSA samples
```{r, message=FALSE}
print(p_combi[, , 1:10])
```

## Other parameters
Parameters for which no uncertainty is propagated in the model
```{r, message=FALSE}
# define LY by health state per cycle
LY <- matrix(c(1, 1, 1, 0),nrow=1)

# define costs by health state per cycle
drug_cost_mono <- matrix(c(2278, 2278, 2278, 0),nrow=1)
drug_cost_combi <- matrix(c(2278+2087, 2278+2087, 2278+2087, 0),nrow=1)
other_cost <- matrix(c(2756, 3052, 9007, 0),nrow=1)
```



# Model output
## Monotherapy
Define array for trace and outcomes by follow-up time for each PSA sample
```{r, message=FALSE}
trace_mono <- array(NA, dim = c(follow_up, 6, PSA_samples))
colnames(trace_mono) <- c("State A", "State B", "State C", "Death", "Disc LY", "Disc cost")
```
Define matrix for expected outcomes for each PSA sample
```{r, message=FALSE}
Expected_LY_mono <- matrix(rep(NA, PSA_samples), nrow=PSA_samples)
Expected_total_cost_mono <- matrix(rep(NA, PSA_samples), nrow=PSA_samples)
```
Trace and expected LY and cost for each PSA sample
```{r, message=FALSE}
for(i in 1:PSA_samples){
for(t in 1:follow_up){
trace_mono[t,1:4,i] <- initial %*% (p_mono[,,i] %^%t)
trace_mono[t,5,i] <- sum(  trace_mono[t,1:4,i] * LY * 1/(1+discount_LY)^t  )
trace_mono[t,6,i] <- sum(  trace_mono[t,1:4,i] * (drug_cost_mono + other_cost) * 1/(1+discount_cost)^t  )
}
Expected_LY_mono[i] <- sum(trace_mono[,5,i])
Expected_total_cost_mono[i] <- sum(trace_mono[,6,i])
}
```
Presentation of results for first 10 PSA samples
```{r, message=FALSE}
trace_mono[,,1:10]
Expected_LY_mono[1:10]
Expected_total_cost_mono[1:10]
```

## Combination therapy
Define array for trace and outcomes by follow-up time for each PSA sample
```{r, message=FALSE}
trace_combi <- array(NA, dim = c(follow_up, 6, PSA_samples))
colnames(trace_combi) <- c("State A", "State B", "State C", "Death", "Disc LY", "Disc cost")
```
Define matrix for expected outcomes for each PSA sample
```{r, message=FALSE}
Expected_LY_combi <- matrix(rep(NA, PSA_samples), nrow=PSA_samples)
Expected_total_cost_combi <- matrix(rep(NA, PSA_samples), nrow=PSA_samples)
```
Trace and expected LY and cost for each PSA sample
```{r, message=FALSE}
for(i in 1:PSA_samples){
for(t in 1:2){
trace_combi[t,1:4,i] <- initial %*% (p_combi[,,i] %^%t)
trace_combi[t,5,i] <- sum(  trace_combi[t,1:4,i] * LY * 1/(1+discount_LY)^t  )
trace_combi[t,6,i] <- sum(  trace_combi[t,1:4,i] * (drug_cost_combi + other_cost) * 1/(1+discount_cost)^t  )
}
for(t in 3:follow_up){
trace_combi[t,1:4,i] <- trace_combi[2,1:4,i] %*% (p_mono[,,i] %^% (t-2))
trace_combi[t,5,i] <- sum(  trace_combi[t,1:4,i] * LY * 1/(1+discount_LY)^t  )
trace_combi[t,6,i] <- sum(  trace_combi[t,1:4,i] * (drug_cost_mono + other_cost) * 1/(1+discount_cost)^t  )
}
Expected_LY_combi[i] <- sum(trace_combi[,5,i])
Expected_total_cost_combi[i] <- sum(trace_combi[,6,i])
}
```
Presentation of results for first 10 PSA samples
```{r, message=FALSE}
trace_combi[,,1:10]
Expected_LY_combi[1:10]
Expected_total_cost_combi[1:10]
```

## Incremental Cost-Effectiveness Ratio
```{r, message=FALSE}
ICER <- matrix(rep(NA, PSA_samples), nrow=PSA_samples)
for(i in 1:PSA_samples){
ICER[i]  <- (Expected_total_cost_combi[i] - Expected_total_cost_mono[i]) / (Expected_LY_combi[i] - Expected_LY_mono[i])
}
```
Presentation of results for first 10 PSA samples
```{r, message=FALSE}
ICER[1:10]
```

## Overview PSA results
```{r, message=FALSE}
Results_by_PSA_sample <- cbind(Expected_LY_mono, Expected_LY_combi, Expected_total_cost_mono, Expected_total_cost_combi, ICER)
colnames(Results_by_PSA_sample) <- c("Disc LY mono", "Disc LY combi", "Disc cost mono", "Disc cost combi", "ICER")
```
Presentation of results first 10 PSA samples
```{r, message=FALSE}
Results_by_PSA_sample[1:10,]
```

# Appendix: Dirichlet distribution
The multinomial distribution is a discrete probability distribution for the number of successes for each of k mutually exclusive categories in n trials. The probabilities of the categories are given by $\pi_1,\ldots, \pi_k$ with $\sum_{j=1}^k \pi_j=1$ and each $\pi_j$ defined on $[0,1]$. The Dirichlet distribution is parameterized by the concentration parameters $\alpha_1,\ldots, \alpha_k$ with $\alpha_j > 0$. Letting $x_1,\ldots, x_k$ denote the number of successes in each category, the prior distribution and likelihood are,

$$
\begin{aligned}
p(\pi_1,\ldots,\pi_k |\alpha_1,\ldots, \alpha_k) = \text{Dirichlet}(\alpha_1,\ldots,\alpha_k) \\
p(x_1,\ldots,x_k | \pi_1,\ldots,\pi_k) = \text{Multin}(n, \pi_1,\ldots,\pi_k).
\end{aligned}
$$